Predicting exercise type by using accelerometers 
============================================================
```{r Setting enviorment, echo=FALSE, results='hide'}
Sys.setlocale("LC_ALL", "C")
Sys.setlocale("LC_MESSAGES", "C")
library(caret)
```


# Synopsis
The goal of the study is to be able to predict exercise type using readings from accelerometers.

This research uses data from [Human Activity Recognition project](http://groupware.les.inf.puc-rio.br/har)

# Data Processing
## Downloading data
At the start data is downloaded and extracted. Alternate way to obtain the data is to download it manually from [link1](https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv) and  [link2](https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv).
```{r Downloading data, results='hide', message=F, warning=F}
require(R.utils)
#download training dataset
download.file("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv", method="curl", destfile="pml-training.csv")
#download testing dataset
download.file("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv", method="curl", destfile="pml-testing.csv")
```

## Loading data
Data is loaded using read.csv function:
```{r Loading data}
pml_training = read.csv("pml-training.csv",
                        stringsAsFactors = FALSE,
                        na.strings = c("","NA"))

pml_testing = read.csv("pml-testing.csv",
                       stringsAsFactors = FALSE,
                       na.strings = c("","NA"))
```

## Preprocessing data
Lets look at the data using str function: 
```{r Data summary, results='hide'}
str(pml_training)
```
There are unnecessary columns that are not accelometers readings nor classe column. Let's remove them:

```{r Event types, results='hide'}
dropCol <- c("X", "user_name", "raw_timestamp_part_1", "raw_timestamp_part_2", 
             "cvtd_timestamp", "new_window", "num_window")
pml_training <- subset(pml_training, select = !names(pml_training) %in% dropCol)
pml_testing <- subset(pml_testing, select = !names(pml_training) %in% dropCol)
```

There are also a lot of columns with more than 75% of NAs values. We will not use this columns in our prediction model:
```{r Remove NAs columns}
classe <- pml_training$classe 
pml_training$classe <- NULL

goodObs <- sapply(pml_training, function(x)sum(!is.na(x))/length(x)<0.25)
n <- names(goodObs)[goodObs==FALSE]

tidy_pml_training <- pml_training[,n]
tidy_pml_testing <- pml_testing[,n]

tidy_pml_training$classe <- as.factor(classe)
```

To check our future assumptions of the out of sample error, we will divide our training data into train and test datasets:

```{r Partitioning data}
set.seed(123)
inTrain <- createDataPartition(y=tidy_pml_training$classe, p=0.8, list=FALSE)
training <- tidy_pml_training[inTrain,]
testing <- tidy_pml_training[-inTrain,]
```

## Model creation
We use random forest classification with 10-fold cross validation to build a machine learning algorithm and make assumptions of the out of sample error.
```{r}
#using 10-fold CV
fitControl <- trainControl(method = "cv", number=10, savePred=T, classProb=T)
RFfit <- train(classe~., data=training, method="rf", trControl = fitControl)
RFfit
```
As we can see, we should have the out of sample error close to zero. Let's check our out of sample error for our testing dataset:
```{r}
#using 10-fold CV
testing$predicted <- predict(RFfit, newdata=testing)
accuracy <- sum(diag(table(testing$predicted, testing$classe)))/nrow(testing)
accuracy
```
Predicting testing dataset yielded ~0.99 accuracy!
Let's create confusion matrix:
```{r}
cm <- confusionMatrix(data=testing$predicted, testing$classe)
cm
```
# Results
Now we can easily predict our pml_testing dataset classe values:
```{r}
answers <- predict(RFfit, newdata=pml_testing)
```
And write them to files:

```{r}
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}

pml_write_files(answers)
```